---
title: "recountmethylation Bioconductor vignette"
author: 
  - name: Sean K. Maden
date: "4/3/2020"
abstract: > 
  This is the main vignette for the `recountmethylation` package. This shows how to obtain and work with HDF5-SummarizedExperiment data, comprising of DNAm assay data and sample metadata from 35,000 samples.
output:
    BiocStyle::html_document:
        toc: true
        tocfloat: true
        code_folding: show
    pdf_document: default
---

# Overview

We piped GEO assay and metadata, available as of March 2019, into large data objects for analysis in R sessions. Data files are versioned and timestamped (see filenames).
The `minfi` and `HDF5Array` libraries were used. 

# obtain data

DNAm assay data and metadata are bundled in H5SE objects. 
These can be downloaded fro (url)
(info about h5se objects)
# access data
Use `HDF5Array` to 
H5SE objects were stored using `saveHDF5SummarizedExperiment`

# work with data
Use `minfi`


# example

First obtain the compact example file from (url). 

```{bash}
url='recount.bio/remethdb_hdf5test_rg'
curl -r $url
```

Next, point to the downloaded file with `loadHDF5SummarizedExperiment`

```{r}
```

Finally, perform data anlaysis tasks as you normally would. For instance:

```{r}
```








