---
title: "recountmethylation Bioconductor vignette"
author: 
  - name: Sean K. Maden
date: "4/3/2020"
abstract: > 
  This is the main vignette for the `recountmethylation` package. This shows how to obtain and work with HDF5-SummarizedExperiment data, comprising of DNAm assay data and sample metadata from 35,000 samples.
output:
    BiocStyle::html_document:
        toc: true
        tocfloat: true
        code_folding: show
    pdf_document: default
---

# Overview

DNAm assay data and sample metadata available as of March 2019 were 
piped into HDF5-based SummarizedExperiment objects roughly 130Gb in size.
These files presently include `#filenames`, located at `#link`

# Download data
Use `HDF5Array` to 
H5SE objects were stored using `saveHDF5SummarizedExperiment`

# Query data

The `HDF5-SummarizedExperiment` objects use `DelayedArray` backend,
which enables rapid querying and delayed execution of memory-intensive processes.
Use `h5se <- HDF5Array::loadHDF5SummarizedExperiment(path)` where `path` is the path to the
downloaded `HDF5-SummarizedExperiment` object. Then access metadata using `minfi::pData(h5se)`.

# Example with test dataset

First obtain the compact example file from `recount.bio/data` with `get_rmdl`.

```{bash}
get_rmdl("h5se-test_gr")
```

Next, point to the downloaded file with `loadHDF5SummarizedExperiment`

```{r}
```

Finally, perform data anlaysis tasks as you normally would. For instance:

```{r}
```








