---
title: "Data Analysis"
author:
- Sean K. Maden
- Reid F. Thompson
- Kasper D. Hansen
- Abhi Nellore
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
  BiocStyle::html_document:
    code_folding: show
    toc: yes
    tocfloat: yes
  html_document:
    df_print: paged
    toc: yes
bibliography: bibliography.bib
package: recountmethylations
vignette: "%\\VignetteDepends{RCurl} %\\VignetteIndexEntry{bioc_vignette}  %\\usepackage[UTF-8]{inputenc}
  %\\VignetteEngine{knitr::knitr} \n"
---

```{r setup, echo = FALSE}
suppressMessages(library(rhdf5))
suppressMessages(library(minfi))
suppressMessages(library(recountmethylation))
suppressMessages(library(knitr))
opts_chunk$set(eval = FALSE, echo = TRUE, 
               warning = FALSE, message = FALSE)
load("data_analyses/env.RData") # load env objects
```

# Overview

Analyses of DNAm array data across studies can pose certain distinct challenges not encounted when performing within-study analysis, not the least of which includes adjustment of study-confounded variables. This vignette walks through two types of analysis that demonstrate how to approach these problems for cross-study compilations accessible using `recountmethylation`.

## Outline

Workflows for two types of analysis are shown. First, blood samples from young and old patients are compared, followed by a comparison of blood and brain tissues. Each analysis includes the following steps.

* Identify samples of interest and define the `sgroup` comparator variable from metadata.
* Generate summary statistics table for sample groups of interest.
* Plot quality control metrics.
* Quality assurance filtering with log2 median methylated and unmethylated signals.
* Normalization and confounder investigation
* Probe set analysis
* Differential methylation analysis

## Data availability

The data will be obtained and loaded. Samples of interest will be subsetted and summarized. Control metrics will be generated for the sample, and samples will be removed based on quality filters.

Download the `RGChannelSet` data with `get_rmdl("\\.h5")` or `get_rmdl("h5se_rg")`, load the data, and store the samples metadata in an object `md` (see `users_guide` vignette for details).

# Analysis 1: Comparison of blood samples from young and old patients

## Sample and group identification and summarization

```{r summary_stats, eval = TRUE}
kable(sst) # table display
```

## Quality control plots

## Quality assurance and filtering

## Normalization and confounder investigation

## Probe set analyses

# Analysis 2: Comparison of blood and brain samples

## Sample and group identification and summarization

```{r summary_stats, eval = TRUE}
kable(sst) # table display
```

## Quality control plots

## Quality assurance and filtering

## Normalization and confounder investigation

## Probe set analyses

# Conclusions

This vignette showed two example anlayses that utilize data objects accessible using `recountmethylation`, with main analysis functions from base R and `minfi`. 

# Session info

```{r get_sessioninfo}
sessionInfo()
```

# Works Cited
