---
title: "Data Analysis"
author:
- Sean K. Maden
- Reid F. Thompson
- Kasper D. Hansen
- Abhi Nellore
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
  BiocStyle::html_document:
    code_folding: show
    toc: yes
    tocfloat: yes
  html_document:
    df_print: paged
    toc: yes
bibliography: bibliography.bib
package: recountmethylations
vignette: "%\\VignetteDepends{RCurl} %\\VignetteIndexEntry{bioc_vignette}  %\\usepackage[UTF-8]{inputenc}
  %\\VignetteEngine{knitr::knitr} \n"
---

```{r setup, echo = FALSE}
suppressMessages(library(HDF5Array))
suppressMessages(library(minfi))
suppressMessages(library(recountmethylation))
suppressMessages(library(knitr))
opts_chunk$set(eval = FALSE, echo = TRUE, 
               warning = FALSE, message = FALSE)
```

# Overview

Analyses of DNAm array data can pose certain distinct challenges not encounted when working with smaller datasets. This vignette shows a minimal (e.g. minimum number of dependencies for plots, etc.) analysis workflow working from the full-sized `RGChannelSet`, an `HDF5-SummarizedExperiment` object using DelayedArray backend.

(summary of sample quantities analyzed)

## Outline

Samples from blood and brain will be identified and characterized.

This includes tables of summary statistics. Certain metadata variables will be used to further filter samples so that only tissue samples (e.g. not cell lines, primary cells, etc.) are included.

Control metrics, including Illumina metrics and log2 methylated/unmethylated signals, will be generated and evaluated. Summaries will be generated for samples groups, studies, etc. Samples with outlying low signals will then be removed. This section will make use of several functions and summarizations that use functions from the `minfi` package (@aryee_minfi:_2014).

Next, noob normalization will be applied. The impact of normalization will be evaluated by comparing Study ID variance contributions before and after noob-norm (@triche_low-level_2013).

# Data summaries and quality assurance

The data will be obtained and loaded. Samples of interest will be subsetted and summarized. Control metrics will be generated for the sample, and samples will be removed based on quality filters.

## Download and load RGChannelSet data

Download the 119Gb dataset with `get_rmdl`, and load the data with the returned path.

```{r}
# path <- get_rmdl("h5se_rg")
library(HDF5Array)
path <- "remethdb_h5se_rg_00-00-01_1583780004"
rg <- loadHDF5SummarizedExperiment(path)
```

## Filter and summarize blood and brain tissue samples

Query the metadata to isolate blood and brain tissue samples using a string match.

```{r}
varname <- "tissue"
mdvar <- mdpost[,varname]
terms <- c("blood", "brain")
strict.regex <- paste0("(^|;)",terms,"($|;)")
cond.match <- grepl(strict.regex, mdvar)|grepl(strict.regex, mdvar)
mdf <- mdpost[,cond.match] # filtered metadata
```

Next obtain key summary statistics to better understand the idnetified samples.

```{r}
sstable <- matrix(ncol = 10)
kable(sstable)
```

To check experiment contexts for matching samples, evaluate the most frequent co-occurring labels as follows.

```{r}
# get labels
tx.labels <- unlist(strsplit(mdf$tissue, ";"))
dx.labels <- unlist(strsplit(mdf$tissue, ";"))
# get label frequencies
dt.tx <- as.data.frame(table(tx.terms))
dt.dx <- as.data.frame(table(dx.terms))
dt.tx <- dt.tx[rev(order(dt.tx[,2])),]
dt.tx <- dt.tx[rev(order(dt.dx[,2])),]
# display tables
kable(dt.tx[c(1:20),])
kable(dt.dx[c(1:20),])
```

## Quality assurance and filters
ctrl_signal()

```{r}
# minfi qc
controlStripPlot()
densityPlot()
```

## Normalization and assessments

## Principal component analysis

## Summary statistics, final

```{r}
kable(sstable)
```

# Differential methylation analyses

## Modeling potential confounders

## T-tests

## F-tests 

# Results summaries

## Enriched regions

## Enriched genes

# Conclusions

# Session info

```{r get_sessioninfo}
sessionInfo()
```

# Works Cited
