---
title: "Data Analysis"
author:
- Sean K. Maden
- Reid F. Thompson
- Kasper D. Hansen
- Abhi Nellore
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
  BiocStyle::html_document:
    code_folding: show
    toc: yes
    tocfloat: yes
  html_document:
    df_print: paged
    toc: yes
bibliography: bibliography.bib
package: recountmethylations
vignette: "%\\VignetteDepends{RCurl} %\\VignetteIndexEntry{bioc_vignette}  %\\usepackage[UTF-8]{inputenc}
  %\\VignetteEngine{knitr::knitr} \n"
---

```{r setup, echo = FALSE}
suppressMessages(library(rhdf5))
suppressMessages(library(minfi))
suppressMessages(library(recountmethylation))
suppressMessages(library(knitr))
opts_chunk$set(eval = FALSE, echo = TRUE, 
               warning = FALSE, message = FALSE)
# load environment data
#load("data_analyses/env.RData")
data("env.RData")
```

# Overview

Analyses of DNAm array data can pose certain distinct challenges not encounted when working with smaller datasets. This vignette shows a minimal (e.g. minimum number of dependencies for plots, etc.) analysis workflow working from the full-sized `RGChannelSet`, an `HDF5-SummarizedExperiment` object using DelayedArray backend.

(summary of sample quantities analyzed)

## Outline

* Samples from blood and brain will be identified and characterized.
* This includes tables of summary statistics. Certain metadata variables will be used to further filter samples so that only tissue samples (e.g. not cell lines, primary cells, etc.) are included.
* Control metrics, including Illumina metrics and log2 methylated/unmethylated signals, will be generated and evaluated. Summaries will be generated for samples groups, studies, etc. Samples with outlying low signals will then be removed. This section will make use of several functions and summarizations that use functions from the `minfi` package (@aryee_minfi:_2014).
* Next, noob normalization will be applied. The impact of normalization will be evaluated by comparing Study ID variance contributions before and after noob-norm (@triche_low-level_2013).

# Data summaries and quality assurance

The data will be obtained and loaded. Samples of interest will be subsetted and summarized. Control metrics will be generated for the sample, and samples will be removed based on quality filters.

## Download and load RGChannelSet data

Download the 119Gb dataset with `get_rmdl`, and load the data with the returned path.

```{r}
path <- "remethdb2.h5"
md <- data_mdpost(path)
```

## Filter and summarize blood and brain tissue samples

This vignette focuses on data from 4 GSE records. For brevity, subsets of these studies will be used.

```{r}
# specify filter values
cn <- "tissue" # colname to query
labs <- c("blood", "brain") # labels of interest
regex.patt <- paste0("(^|;)", labs,"(;|$)") # pattern for strict label matches
# define and filter on sgroup variable
mdf$sgroup <- "NA"
sg.blood <- grepl(regex.patt[1], mdf[,cn]) & 
      !grepl(regex.patt[2], mdf[,cn])
sg.brain <- grepl(regex.patt[2], mdf[,cn]) & 
      !grepl(regex.patt[1], mdf[,cn])
mdf[sg.blood,]$sgroup <- labs[1]
mdf[sg.brain,]$sgroup <- labs[2]
mdf <- mdf[sg.blood|sg.brain,]
# select 50 samples per groups
set.seed(2)
ngsm.sg <- 50
gsm.blood <- which(mdf$sgroup == labs[1])
gsm.brain <- which(mdf$sgroup == labs[2])
gsm.indices <- c(sample(gsm.blood, ngsm.sg), sample(gsm.brain, ngsm.sg))
mdf <- mdf[gsm.indices,]
```

Next, generate a summary of key statistics for each sample group of interest. First, define some function `get_sst` that returns a neat summary statistics table. For comparison, `get_sst` will be rerun after quality assessments and filters have been applied.

```{r}
get_sst <- function(labs = c("blood", "brain")){
  # get summary stats by label
  sstat <- lapply(labs, function(x){
    mv <- mdf[mdf$sgroup==x,]
    v1 <- nrow(mv)
    numgsm.gse <- as.data.frame(table(mv$gseid))
    v2 <- round(mean(numgsm.gse[,2]), 3)
    v3 <- round(sd(numgsm.gse[,2]), 3)
    v4 <- nrow(numgsm.gse)
    # predage summaries
    pv <- as.numeric(mv$predage)
    v5 <- min(pv)
    v6 = max(pv)
    v7 <- round(mean(pv, na.rm = TRUE), 3)
    v8 <- round(sd(pv, na.rm = TRUE), 3)
    v9 <- length(pv[is.na(pv)])
    # predsex summaries
    v10 <- round(100*nrow(mv[mv$predsex=="F",])/nrow(mv), 3) # perc. F
    v11 <- nrow(mv[is.na(mv$predsex),])
    return(data.frame("ngsm" = v1, "meangsm.gse" = v2, "sdgsm.gse" = v3,  "numgse" = v4,
                    "min.predage" = v5, "max.predage" = v6, "mean.predage" = v7, 
                    "sd.predage" = v8, "numna.predage" = v9, "percfemale.predsex" = v10, 
                    "numna.predsex" = v11))
  })
  sst <- sapply(sstat, rbind) # bind label groups
  colnames(sst) <- labs
  return(sst)
}
sst1 <- get_sst()
save.image(file = "data_analyses/env.RData")
```

The following tall table was generated.

```{r summary_stats, eval = TRUE}
kable(sst) # table display
```

## Quality assurance and filters
ctrl_signal()

```{r}
# minfi qc
controlStripPlot()
densityPlot()
```

## Normalization and assessments

## Principal component analysis

## Summary statistics, final

```{r}
kable(sstable)
```

# Differential methylation analyses

## Modeling potential confounders

## T-tests

## F-tests 

# Results summaries

## Enriched regions

## Enriched genes

# Conclusions

# Session info

```{r get_sessioninfo}
sessionInfo()
```

# Works Cited
