---
title: "recountmethylation Bioconductor vignette"
author: 
  - name: Sean K. Maden
  - name: Reid F. Thompson
  - name: Kasper D. Hansen
  - name: Abhi Nellore
package: recountmethylation
date: "4/4/2020"
abstract: > 
  This is the main vignette for the `recountmethylation` package.
vignette: >
    %\VignetteDepends{RCurl}
    %\VignetteEngine{knitr::knitr} 
    %\VignetteIndexEntry{bioc_vignette} 
    %\usepackage[UTF-8]{inputenc}
output:
    BiocStyle::html_document:
        toc: true
        tocfloat: true
        code_folding: show
    pdf_document: default
---

```{r setup, include = FALSE, eval = TRUE}
library(recountmethylation)
knitr::opts_chunk$set(eval = TRUE,
    echo = TRUE, warning = FALSE,
    message = FALSE)
```

# Overview

DNAm assay data and sample metadata available as of March 2019 were 
piped into HDF5-based SummarizedExperiment objects roughly 130Gb in size.
These files presently include `#filenames`, located at `#link`

# Download data

DNAm assay and sample metadata is stored at `recount.bio/data`. 
We recommend working with data formatted as `HDF5SummarizedExperiment`, including
`remethdb_h5se_gm_00-00-01_1583780004` (`GenomicMethylSet` format),
`remethdb_h5se_gr_00-00-01_1583780004` (`GenomicRanges` format), 
`remethdb_h5se_rg_00-00-01_1583780004` (`RGChannelSet` format). 
Note these files are relatively large (about 80 - 120 Gb). 
However, these make use of `DelayedArray` and `HDF5Array` to 
allow rapid data queries, summaries, and management of preprocessing workflows.
Download these from your browser or using the `get_rmdl()` function in `recountmethylation`.

# Query data

Use `h5se <- HDF5Array::loadHDF5SummarizedExperiment(path)` where `path` is the path to the
downloaded `HDF5-SummarizedExperiment` object. Then access metadata using `minfi::pData(h5se)`. See the minfi [page](https://bioconductor.org/packages/release/bioc/html/minfi.html) for more details.

# Example with test dataset

First obtain the test data from `recount.bio/data` with `get_rmdl()`, which returns the path we'll use to load the data.

```{r, testdat.dl}
datapath <- get_rmdl("h5se-test_gr", verbose = FALSE)
```

Once data is successfully downloaded, use the returned path to load data into an R session with `HDF5Array::loadHDF5SummarizedExperiment()` thusly:

```{r, testdat.load}
grtest <- HDF5Array::loadHDF5SummarizedExperiment(datapath)
dim(grtest)
```

Data summaries, analysis, and preprocessing can be performed, such as with `minfi`.

# Session info
```{r}
sessionInfo()
```
